Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-25 article/main.tex
modifications mineurs

\documentclass[preprint,svgnames]{iacrtrans}

@@

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' it is passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites (\todo{les
  mentionner}). The goal of designers then consists in designing the fastest
possible generator that passes the day's favorite test suite.

@@

The internal state of the \textsf{PCG64} generator is made of a 128-bit
``state'' and a 128-bit ``increment'', whose intended use is to provide several
pseudo-random streams with the same seed (just as the initialization vectors do
in stream ciphers). A default increment is provided in case the end-user just
want one pseudo-random stream with a single 128-bit seed.

@@

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialized using an external source of entropy. In this case, our prediction
algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves and instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

@@

Figure~\ref{pcg128out} summarizes the process. The overall design strategy is
similar to that of a filtered LFSR: the successive states of a weak internal
generator with a strong algebraic structure are ``filtered'' by a non-linear
function.

@@

\subsection{A first way to find U : the Babai's rounding}

We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\).

@@

If we can prove that the right side \todo{je confonds ma gauche et ma droite et je vous zut :p} of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.

 \subsection{A second way to find U : using an exact CVP solver}

Solving exactly the CVP is a hard computational problem. But as we do not manipulate quantities too big, we still can use an exact CVP solver.

We know that \(U\) is in the lattice, \(T\) is not and \(|T-U|\) is small. Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).

@@

We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\). We notice that \(U\) is in \(L(A,2^{64},n)\).

@@

>>>>

% !TeX spellcheck = en
\documentclass[preprint,svgnames]{iacrtrans}

@@

In the realm of non-cryptographic random generators, a PRG is deemed ``good
enough'' it is passes \emph{some} efficient statistical tests --- whereas the
cryptographic notion of pseudo-randomness asks that it passes \emph{any}
efficient test. There are \textit{de facto} statistical test suites (\todo{les
  mentionner}). The goal of designers then consists in designing the fastest
possible generator that passes the day's favourite test suite.

@@

The internal state of the \textsf{PCG64} generator is made of a 128-bit
``state'' and a 128-bit ``increment'', whose intended use is to provide several
pseudo-random streams with the same seed (just as the initialisation vectors do
in stream ciphers). A default increment is provided in case the end-user just
want one pseudo-random stream with a single 128-bit seed.

@@

When the increment is unknown, things are a bit more complicated. This is the
default situation in \textsf{NumPy}, where both the state and the increment are
initialised using an external source of entropy. In this case, our prediction
algorithm requires 3072 bits of pseudo-random stream ; it guesses between 51 and
55 bits, then for each guess it solves and instance of CVP in dimension 4 (using
about 75 arithmetic operations). This recovers 64 more bits of information about
the difference between two successive states, and this is enough to filter the
bad guesses. This information can then be used in a subsequent and comparably
inexpensive phase to recover the entire internal state. On average, the whole
process requires a bit less than 20 000 CPU hours to complete.

@@

Figure~\ref{pcg128out} summaries the process. The overall design strategy is
similar to that of a filtered LFSR: the successive states of a weak internal
generator with a strong algebraic structure are ``filtered'' by a non-linear
function.

@@

 \subsection{A first way to find U : using an exact CVP solver}

Solving exactly the CVP is a hard computational problem. The algorithms used are costly, but we can afford to use them up to dimension 50.

We know that \(U\) is in the lattice, \(T\) is usually not and \(|T-U|\) is small. Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).

\subsection{A second way to find U : the Babai's rounding}

As using an exact CVP solver is costly, we'll try a cheaper but less effective way to find the closest vector.

We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Ax \rVert_2}{\lVert x \rVert_2}\]
In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).

\medskip

@@

If we can prove that the right side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.

We see that  the right side of the inequality is often larger than in the first case. We will need to use larger parameters to find the vector we are looking for. But as the Babai rounding is largely cheaper than using an exact CVP solver, there will be cases where this method is far more practical. \todo{je trouve la formulation un peu moche}

@@

The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero. As \(S'\) is in \(L(A,2^{128},n)\), \(S' \mod [2^{64+\ell}]\) is in \(L(A,2^{64+\ell},n)\)

We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\). As the \(\ell\) lower bits of \(S'\) are zero, we have \(S' \equiv U \times 2^\ell\ \mod [2^{64+\ell}]\). Hence \(U\) is in \(L(A,2^{64},n)\).

@@

\todo{maybe dessin}
